---
title: "EPIDEMIOLOGY OF MULTI-DRUG RESISTANT TB USING ISOLATES SENT TO THE NATIONAL TUBERCULOSIS REFERENCE LABORARY (2015-2016)"
format:
  pdf:
    documentclass: report
    keep-tex: true
---

# METHODS

```{R}
library('ggflowchart')
flow_chart <- tibble::tibble(from = c("Total TB+ Entries recieved 
  8308", "Total TB+ Entries recieved 
  8308", "Entries after removing null values
  2621", "Entries after removing null values
  2621", "Null entries removed
  2621"),
                             to= c("Entries after removing null values
  2621","Entries removed 
  5682", "Entries after removing Non-MDR
  143
  ", "Null entries removed
  2621", "MDR entries analyzed
  143"))
ggflowchart(flow_chart, colour = 'cyan', family = 'serif', x_nudge = 0.45)

```

## R Packages used

```{r}
#\ label: load-packages
suppressPackageStartupMessages(library('tidyverse'))

#install.packages('plotly')

#install.packages('ggmap')

#install.packages('forcats')
suppressPackageStartupMessages(library('ggmap'))
suppressPackageStartupMessages(library('plotly'))
suppressPackageStartupMessages(library('ggdist'))
suppressPackageStartupMessages(library('ggmap'))

```

# FUNCTIONS

## Fisher's function

```{r}
Fishersfn <- function(df, var1, var2){

  contingency_table <- table(df[[var1]], df[[var2]])

  fisherExact <- fisher.test(contingency_table)

  return (fisherExact)

}
```

## Getting percentages

```{r}
pcts <- function(df, var_to_group_by) {

  results <- df %>%

    group_by_at(vars({{ var_to_group_by }})) %>%

    summarise(counts = n()) %>%

    mutate(Percentages = counts/sum(counts)*100)

  

  return(results)

}
```

Reading the file

```{r}
mdr_tb_df <- read.csv('mdr_tbv4.csv')

```

Creating a new column to define MDR_TYPE

```{r}
mdr_tb_df$MDR_TYPE <- NA
```

Here I have defined Rpob + katg as MDRKR (MDR KatG and RpoB), as well as those that are rpob + katg + inha

```{r}
mdr_tb_dfv1 <- mdr_tb_df$MDR_TYPE[mdr_tb_df$RpoB == 'Resistant' & mdr_tb_df$KatG == 'Resistant'] <- 'MDRKR'

mdr_tb_df$MDR_TYPE[mdr_tb_df$RpoB == 'Resistant' & mdr_tb_df$KatG != 'Resistant' & mdr_tb_df$inhA == 'Resistant'] <- 'MDRIR'

mdr_tb_dfv1 <- mdr_tb_df %>% select(-RpoB, -KatG, -inhA)
```

# VISUALIZATIONS

Setting theme

```{r}
theme_set(theme_gray())

theme_update(

  plot.margin = margin(rep(20, 4))

)

```

## HIV BAR PLOT

```{r}
hivbar <- ggplot(mdr_tb_dfv1, mapping = aes(x=HIV_STATUS)) + geom_bar(mapping = aes(fill = HIV_STATUS)) + ggtitle('HIV status and multidrug resistance in 2015-2016 samples from NTRL')
hivbar
```

## TB TYPE BAR PLOT

```{r}
tb_type_bar <- ggplot(mdr_tb_dfv1, mapping = aes(x=TB_TYPE)) + geom_bar(mapping = aes(fill = TB_TYPE)) + ggtitle('TB Type and multidrug resistance
                                                                                                                 in 2015-2016 samples from NTRL')
```

## SMEAR BAR PLOT

```{r}
smearBar <- ggplot(mdr_tb_dfv1, mapping = aes(x=SMEAR)) + geom_bar(mapping = aes(fill = SMEAR)) + ggtitle('Smear results and multidrug resistance in 2015-2016 samples from NTRL')
smearBar 
```

Age groups based on WHO typical age divisions

```{r}
mdr_tb_dfv3 <- mdr_tb_dfv1 %>%   mutate(
  # Create categories
  age_group = dplyr::case_when(
    AGE <= 4            ~ "0-4",
    AGE > 4 & AGE <= 14 ~ "5-14",
    AGE > 14 & AGE<= 24 ~ "15-24",
    AGE > 24 & AGE <= 34  ~ "25-34",
    AGE > 34 & AGE <= 44 ~ "35-44",
    AGE > 44 & AGE <= 54 ~ "45-54",
    AGE > 54 & AGE <= 64 ~ "65+",
    AGE >= 64 ~ "35-44"
  ),
  # Convert to factor
  age_group = factor(
    age_group,
    level = c("0-4", "5-14","15-24","25-34","35-44","45-54", "65+")
  )
)
```

## LOLLIPOP PLOT FOR COUNTY/GEOGRAPHICAL DISTRIBUTION

```{r}
geoGrpsPcts <- pcts(mdr_tb_dfv1, 'Province')

GeoPops <- geoGrpsPcts %>%

  arrange(desc(Percentages)) %>% 

  mutate( Province = forcats::fct_reorder(Province, Percentages)) %>%

  ggplot(aes(x=Province, y=Percentages)) +

  geom_point(color = 'orange') +

  geom_segment(aes(x=Province, y=Percentages, xend=Province, yend=0, color = 'red')) +

  theme(axis.text.x = element_text(angle = 90), panel.background = element_rect(fill = 'lightblue'), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), plot.title = element_text(color='#6699CC', size = 10, face = 'bold')) +

  labs(title = 'MDR TB: Percentage of MDR TB cases by region/county reported to the NTRL', y= 'Percentage (%) per county', x='County') +

  guides(color = 'none')
GeoPops 
```

## AGE GROUP BAR PLOT

```{r}
age_groupBars <- mdr_tb_dfv3 %>%

  group_by(age_group) %>%

  summarise(counts = n())

ggplot() +

  geom_bar(data = age_groupBars, aes(x = age_group, y = counts, fill = age_group), stat = "identity") +

  labs(title = 'Age distribution of MDR isolates subitted to the NTRL between 2015-2015')
age_groupBars
```

# **INFERENTIAL STATISTICS**

## Fishers exact test for HIV status

```{r}
p_value <- Fishersfn(mdr_tb_dfv3, 'HIV_STATUS', 'MDR_TYPE')
p_value 
```

## Fisher's exact test for age

```{r}
p_value <- Fishersfn(mdr_tb_dfv3, 'age_group', 'MDR_TYPE')
p_value 
```

## Fisher's exact test for COUNTY

```{r}
p_value <- Fishersfn(mdr_tb_dfv1, 'Province', 'MDR_TYPE')
p_value
```

## Fisher's exact test for TB type

```{r}
p_value <- Fishersfn(mdr_tb_dfv1, 'TB_TYPE', 'MDR_TYPE')
p_value
```

## Fisher's exact test for GENDER

```{r}
p_value <- Fishersfn(mdr_tb_dfv1, 'GENDER', 'MDR_TYPE')
p_value
```

## Fisher's exact test for SMEARS

```{r}
p_value <- Fishersfn(mdr_tb_dfv1, 'SMEAR', 'MDR_TYPE')
p_value
```

# More characterizations

```{r}
mdr_tb_dfv5 <- mdr_tb_df$MDR_TYPE[mdr_tb_df$RpoB == 'Resistant' & mdr_tb_df$KatG == 'Resistant' & mdr_tb_df$inhA == 'Resistant'] <- 'MDR+'
```

# Percentages of the MDR types

```{r}
pcts(mdr_tb_df, 'MDR_TYPE')
```

TABLES OF DEMOGRAPHIC DISTRIBUTION OF MDR TYPES

```{r}
table(mdr_tb_df$Province, mdr_tb_df$MDR_TYPE)

table(mdr_tb_df$HIV_STATUS, mdr_tb_df$MDR_TYPE)

table(mdr_tb_df$SMEAR, mdr_tb_df$MDR_TYPE)

table(mdr_tb_df$TB_TYPE, mdr_tb_df$MDR_TYPE)


```
